---
title: "EC2 CLC Data Analysis"
author: "AMP"
date: "`r Sys.Date()`"
output: html_document
---

Coastal Land Conversion Project

Rationale: to look at more localized drivers of VRR based on Molino et al., 2023 hypothesis that these are dependent on local drivers. https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2023JG007525

Datasets:

1) vertical retreat rate

Dataset from: Chen and Kirwan 2023 https://onlinelibrary.wiley.com/doi/10.1111/gcb.17081 
- vertical forest retreat rates for every point along the mid-Atlantic https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-vcr.399.1


2) threshold elevations

Dataset from: Molino et al 2022 https://aslopubs.onlinelibrary.wiley.com/doi/10.1002/lol2.10262 
- threshold elevation defining the boundary between forest and marsh, aggregated by watershed. This is only for Maryland and Virginia, so won’t work for EXCHANGE sites in DE or NJ
https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-vcr.343.2

3) Watershed parameters from Molino et al 2022 

4) Soil parameters from EC1 v3 

https://data.ess-dive.lbl.gov/view/doi%3A10.15485%2F1960313


Analysis to be completed in this script:

1) Multiple linear regression of vertical retreat rate vs. EC1 soil parameters 

2) Multiple linear regression of threshold elevations vs. EC1 soil parameters (+identified watershed parameters from Molino et al)

```{r Import necessary packages}
library(readr)
library(dplyr)
library(ggplot2)
library(car)
library(sf)
library(raster)
library(tidyverse)
```

### Load in all datasets: 

```{r data loading, warning=FALSE, echo=FALSE}
#1. vertical Forest Retreat Rate
vfrr_shp <- st_read("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/knb-lter-vcr.399.1/Vertical_forest_retreat_rate/Vertical_forest_retreat_rate.shp")

#2. threshold elevations 

elevation_df <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/knb-lter-vcr.343.2/CB_landuse.csv", skip = 23)

HUC_shp <- st_read("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/knb-lter-vcr.343.2/CB_HUC10/CB_HUC10.shp")
```

```{r Import csv file with HUC10 watershed-level variables. There are 14 variables total.}
HUC14var <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/knb-lter-vcr.387.1/VCR23383.csv", skip = 23) 

colnames(HUC14var) <- c('FID', 'Watershed', 'Elevation', 'Tidal_Range', 'Surface_Salinity',  'Average_Slope', 
                         'Isabel_maximum_inundation_depth', 'Isabel_Inundation_duration', 'Marsh_width', 'Watershed_area', 
                         'TPI', 'Annual_temperature', 'Annual_precipitation', 'Annual_growing_season', 'WaterOccurrence', 
                         'Change_WaterOccurrence', 'NDWI')
head(HUC14var)
```
For WRT:
α is a scale parameter that is inversely proportional to the mean pore diameter; n is a shape parameter that controls the slope of the soil-water characteristic curve; θr is the residual water content, and θs is the saturated water content.

```{r import ec1 data}
#4. EC 1 Metadata
ec1_meta <-read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_metadata_v3/ec1_metadata_collectionlevel.csv")

ec1_meta_kit <-read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_metadata_v3/ec1_metadata_kitlevel.csv")

#EC 1 Soil Data
ec1_bd <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_bulk_density_L2.csv") %>%
  dplyr::select(-igsn)
ec1_cond <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_cond_L2.csv") %>%
    dplyr::select(-igsn)
ec1_gwc <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_gwc_L2.csv") %>%
    dplyr::select(-igsn)
ec1_ph <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_ph_L2.csv") %>%
    dplyr::select(-igsn)
ec1_mineralogy <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_mineralogy_L2.csv") %>%
    dplyr::select(-igsn)
ec1_texture <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_texture_L2.csv") %>%
   dplyr::select(-igsn)
ec1_wrc<- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_soil_v3/ec1_soil_wrc_L2.csv") %>%
  dplyr::select(-igsn, -th_r, -n)


#EC 1 adjacent surface water data:
ec1_water_sal <- read_csv("~/GitHub/EC2-Coastal-Land-Conversion/data_do_not_commit/EC1 Package V3/ec1_water_v3/ec1_water_salinity_L2.csv") %>%
  rename(surface_water_sal_psu = sal_psu) %>%
  dplyr::select(kit_id, surface_water_sal_psu)

```

### Clean and merge datasets 

#### EXCHANGE datasets and metadata

```{r cleaning data, warnings=FALSE, echo=FALSE}
ec1_meta_curated <- ec1_meta %>%
  full_join(ec1_meta_kit, by="kit_id") %>%
  filter(region == "Mid-Atlantic") %>%
  #can't do this analysis on the wetland, sediment or water data
  #Want to look for the variables for the upland, though
  filter(transect_location %in% c("upland")) %>%
  filter(!is.na(latitude)) %>%
  dplyr::select(kit_id, region, transect_location, site_name, latitude, longitude, elevation_m, soil_type, soil_horizon)

ec1_meta_curated

ec1_soil_all <- ec1_bd %>%
  full_join(ec1_cond, by=c("campaign", "kit_id", "transect_location")) %>%
  full_join(ec1_gwc, by=c("campaign", "kit_id", "transect_location")) %>%
  full_join(ec1_ph, by=c("campaign", "kit_id", "transect_location")) %>%
  full_join(ec1_mineralogy, by=c("campaign", "kit_id", "transect_location")) %>%
  full_join(ec1_texture, by=c("campaign", "kit_id", "transect_location")) %>%
  full_join(ec1_wrc, by=c("campaign", "kit_id", "transect_location")) %>%
  inner_join(ec1_meta_curated, by=c("kit_id","transect_location"))

ec1_soil_all
```


#### Vertical Forest Retreat Rate

Projected Coordinate Reference System (CRS) "WGS 84 / UTM zone 18N" = EPSG code for this CRS is 32618.

```{r data extracting, echo=FALSE}

# Focusing on upland sites for lat and long: 
coordinates <- ec1_meta_curated %>%
  dplyr::select(kit_id, transect_location, longitude, latitude) %>% 
   filter(!(kit_id == "K055" & transect_location == "upland"),
          !(kit_id == "K058" & transect_location == "upland"),
          !(kit_id == "K042" & transect_location == "upland"),
         !(kit_id == "K042" & transect_location == "transition"),
          !(kit_id == "K045" & transect_location == "upland"),
         !(kit_id == "K045" & transect_location == "transition"),
          !(kit_id == "K048" & transect_location == "upland"),
         !(kit_id == "K048" & transect_location == "transition")
  ) # no retreat data here

coordinates_sf <- st_as_sf(coordinates, coords = c("longitude", "latitude"), crs = 4326)

# Transform the CRS to match the shapefile
coordinates_sf_transformed <- st_transform(coordinates_sf, 32618)

extracted_vfrr_data <- st_join(coordinates_sf_transformed, vfrr_shp, join = st_intersects)

extracted_vfrr_no_geometry <- st_set_geometry(extracted_vfrr_data, NULL)

extracted_vfrr_df <- as.data.frame(extracted_vfrr_no_geometry) %>%
  #filter out the -9999 values
  filter(V_RetreatR > 0 & V_RetreatR < 99) %>%
  #filter out places with NAs
  filter(!is.na(V_RetreatR))

```

```{r data merging, echo=FALSE}

clc_all_data <- extracted_vfrr_df %>%
  inner_join(ec1_soil_all, by= c("kit_id","transect_location")) %>%
  left_join(ec1_water_sal, by = "kit_id")
```

#### Threshold Elevations 

Projected CRS: NAD83 / UTM zone 18N = crs epsg 26918 aka should be the same as  st_crs(HUC_shp)
```{r data extracting in other crs, echo=FALSE}

ec1_meta_curated_2 <- ec1_meta %>%
  full_join(ec1_meta_kit, by="kit_id") %>%
  filter(region == "Mid-Atlantic") %>%
  #can't do this analysis on the wetland, sediment or water data
  filter(transect_location %in% c("upland")) %>%
  filter(!is.na(latitude)) %>%
  dplyr::select(kit_id, region, transect_location, site_name, latitude, longitude, elevation_m, soil_type, soil_horizon)

coordinates_ws <- ec1_meta_curated_2 %>%
  dplyr::select(kit_id, transect_location, longitude, latitude) 

coordinates_sf_1 <- st_as_sf(coordinates_ws, coords = c("longitude", "latitude"), crs = 4326)

# Transform the CRS to match the shapefile
coordinates_ws_sf_transformed <- st_transform(coordinates_sf_1, 26918)


points_in_watershed <- st_join(coordinates_ws_sf_transformed, HUC_shp, join = st_within) %>%
  drop_na()



```

```{r data merging elevation, echo=FALSE}

points_in_watershed_no_geometry <- st_set_geometry(points_in_watershed, NULL)

clc_all_data_ws <- points_in_watershed_no_geometry %>%
  inner_join(ec1_soil_all, by= c("kit_id", "transect_location")) %>%
  left_join(ec1_water_sal, by = "kit_id")

clc_all_data_ws 

```

```{r merging in huc level data w elevation, echo=FALSE}

HUC14var_sig <- HUC14var %>%
  dplyr::select(Watershed, Tidal_Range, Surface_Salinity, Average_Slope) %>%
  rename(name = Watershed)

clc_all_data_ws_sigcors <- clc_all_data_ws %>%
  left_join(HUC14var_sig, by = "name")

```
missing HUC14 data for: 
- Upper Patuxent River
- Huntington Creek-Choptank River

Need to confirm with Matt/Grace if we can substitute others in or not.

### Perform MLRs. 

#### create auto-stepwise function 

```{r autostepwise}

auto_stepwise_regression <- function(data, response_variable, vif_threshold, corr_threshold, direction = "both", significance_level = 0.05) {
  # Helper function to calculate VIF and remove high VIF columns
  remove_high_vif <- function(model, data, vif_threshold) {
    vif_vals <- vif(model)
    while (any(vif_vals > vif_threshold)) {
      highest_vif_var <- names(vif_vals)[which.max(vif_vals)]
      data <- data[ , !(names(data) %in% highest_vif_var)]
      model <- lm(as.formula(paste(response_variable, "~ .")), data = data)
      vif_vals <- vif(model)
    }
    return(list(model = model, data = data))
  }
  
  # Helper function to remove highly co-correlated variables
  remove_high_corr <- function(data, corr_threshold) {
    corr_matrix <- cor(data, method = "pearson") # Uses pairs only when we have complete observations 
    drop_vars <- findCorrelation(corr_matrix, cutoff = corr_threshold, exact = FALSE)
    data <- data[ , -drop_vars, drop = FALSE]
    
    # Print the correlation matrix using corrplot
    library(corrplot) # Needed for the correlation plot
    corrplot(corr_matrix, method = "color", type = "lower", tl.col = "black", tl.srt = 45, title = sprintf("Correlation Matrix (Threshold: %s)", corr_threshold))
    
    # Print the columns that are kept
    print("Columns kept after removing high co-correlation:")
    print(colnames(data))
    return(data)
  }
  
  # Load necessary libraries
  library(caret) # Needed for findCorrelation function
  library(car) # Needed for VIF function, vif
  
  # Prepare the initial dataset excluding the response variable
  initial_data <- data[, !colnames(data) %in% response_variable]
  
  # Step 1: Remove high co-correlation variables
  cleaned_data <- remove_high_corr(initial_data, corr_threshold)
  
  print("data after high correlations removed")
  print(cleaned_data)
  
  # Step 2: Add the response variable back
  cleaned_data[[response_variable]] <- data[[response_variable]]
  
  # Step 4: Prepare the full model using the cleaned data
  formula <- as.formula(paste(response_variable, "~ ."))
  full_model <- lm(formula, data = cleaned_data)
  
  summary(full_model)
  
  # Step 5: Remove high VIF variables
  result <- remove_high_vif(full_model, cleaned_data, vif_threshold)
  cleaned_data <- result$data
  
  # Step 6: Define null and updated full model for stepwise regression
  null_model <- lm(as.formula(paste(response_variable, "~ 1")), data = cleaned_data)
  full_model <- lm(as.formula(paste(response_variable, "~ .")), data = cleaned_data)
  
  # Step 7: Perform stepwise regression based on p-values
  stepwise_model <- step(full_model, 
                         scope = list(lower = null_model, upper = full_model), 
                         direction = direction, 
                         test = "F", 
                         keep = function(model, data) {
                           significant_vars <- summary(model)$coefficients[, "Pr(>|t|)"] < significance_level
                           return(significant_vars)
                         },
                         trace = 2)
  
  # Step 8: Return the final stepwise model
  return(stepwise_model)
}

```

#### Run MLR for Vertical Retreat Rates

```{r vrr mlr}
#clean up dataframe to numerical variables of interest: 
clc_for_vrr_model <- clc_all_data %>%
  dplyr::select(V_RetreatR, bulk_density_g_cm3:albite, kaolinite, microcline, quartz:percent_silt, elevation_m, surface_water_sal_psu) %>% 
  mutate_if(is.numeric, function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x))

clc_for_vrr_model 

VRR_model <- auto_stepwise_regression(clc_for_vrr_model, "V_RetreatR", vif_threshold = 10, corr_threshold = 0.5, direction = "both")

# Show summary of the final model
summary(VRR_model)

#Variable importance of the fit model: 
VRR_importance <- varImp(VRR_model, scale = TRUE)

# Convert to data frame for ggplot2
VRR_importance_df <- as.data.frame(VRR_importance)
VRR_importance_df$Variable <- rownames(VRR_importance_df)

# Create a plot
ggplot(VRR_importance_df, aes(x = reorder(Variable, Overall), y = Overall)) +
    geom_bar(stat = "identity") +
    coord_flip() + 
    xlab("Variables") + 
    ylab("Importance (scaled)") 

```

#### Run MLR for Threshold Elevation 

```{r vrr mlr}
#clean up dataframe to numerical variables of interest: 
clc_for_te_model <- clc_all_data_ws_sigcors %>%
  dplyr::select(Elevation, bulk_density_g_cm3:albite, microcline, quartz:th_s, surface_water_sal_psu:Average_Slope) %>%   
  mutate_if(is.numeric, function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x))

clc_for_te_model

TE_model <- auto_stepwise_regression(clc_for_te_model , "Elevation", vif_threshold = 5, corr_threshold = 0.5, direction = "both")

# Show summary of the final model
summary(TE_model)

```

```{r plotting} 
#Variable importance of the fit model: 
TE_importance <- varImp(TE_model, scale = TRUE)

# Convert to data frame for ggplot2
TE_importance_df <- as.data.frame(TE_importance)
TE_importance_df$Variable <- rownames(TE_importance_df)

#clean up names for plotting:
TE_importance_df <- TE_importance_df %>%
mutate(Variable = case_when(Variable == "percent_clay" ~ "% clay",
                            Variable == "th_s" ~ "saturated water content",
                            Variable == "alpha" ~ "porosity parameter (alpha)",
                            Variable == "Tidal_Range" ~ "tidal range in HUC 10 watershed",
                            Variable == "ph" ~ "soil ph",
                            Variable == "albite" ~ "albite content",
                            Variable == "microcline" ~ "microcline content", 
                            TRUE ~ Variable))

# Create a plot
ggplot(TE_importance_df, aes(x = reorder(Variable, Overall), y = Overall, fill = Variable)) +
    geom_bar(stat = "identity") +
    coord_flip() + 
     scale_fill_manual(values=PNWColors::pnw_palette(name ="Lake", 7, type = "discrete"))+
    xlab("Significant Variables") + 
    ylab("Scaled Importance") +
    theme_classic() +
    theme(legend.position = "none")
    
```

α is a scale parameter that is inversely proportional to the mean pore diameter; n is a shape parameter that controls the slope of the soil-water characteristic curve; θr is the residual water content, and θs is the saturated water content.
